---
title: "Notebook 1"
author: "your name"
date: "June 22, 2022"
output: html_notebook
---

# Fundamentals of Hydrology: Streamflow

Today, we are going to explore stream flow data. These data span several decades and are part of a Long Term Ecological Research Forest in Oregon. The [H.J.Andrews LTER](https://andrewsforest.oregonstate.edu/) is located in the Oregon Cascades. But before looking at these data, we need to install some packages that are going to make it easier for us


## Loading packages

Once we have installed the packages, we need to load them into our workspace, so they can do their magic for us. Let's start with a great one: ggplot2. [ggplot2](https://ggplot2.tidyverse.org/) is one of the most complete packages for data visualization, if you are curios about more details, take a look at the webpage. To load a package we will use the command "library". You need to do this just once, so I normally load my packages at the begining of my session in r.

```{r}
library(ggplot2)
```

## Loading the data

A detailed explanation of these datasets can be found at the [H.J.Andrews LTER](https://andrewsforest.oregonstate.edu/) webpage. More context about our study case can be found in this [MURAL](https://app.mural.co/t/oregonstate7865/m/oregonstate7865/1656115879758/38f3b23d60a92c45a68ecd8ac9b59572fd9dbc6e?sender=uaa06939fbabfc43444e21792) for one of my recent scientific presentations. We will go over this study case in class.

```{r}
#Loading the data and making some internal arrangement
hjg <- read.csv("220621_fho_hjandrews.csv")
hjg$ws.f <- factor(hjg$ws.f,levels = c("Old-growth","Logged"))
hjg$ssn <- factor(hjg$ssn,levels = c("Fall","Winter","Spring","Summer"))

#Defining a common theme for our graphics
theme_fg = theme(axis.text=element_text(colour="black",size=14),
                 axis.title.y = element_text(size = 20),
                 axis.title.x = element_text(size = 20),
                 panel.grid.minor= element_line(colour = "gray", linetype = "dotted"), 
                 panel.grid.major = element_line(colour = "gray", linetype = "dashed"),
                 panel.border = element_rect(fill=NA, colour = "black", size=1),
                 panel.background=element_blank(),
                 axis.ticks.length = unit(0.254, "cm"),
                 axis.ticks = element_line(colour = "black", size=1), 
                 axis.line = element_line(colour = "black"),
                 legend.position = c(.90,.90),
                 legend.direction = "vertical",
                 legend.background = element_rect(fill=alpha(0.1)),
                 legend.title = element_blank())
```



Let's take a look at the raw discharge data.

```{r}
uq_plot <- ggplot(hjg, aes(y=uq.cm, x=wd,color=as.factor(ws.f)))+
  geom_line(alpha=0.6, size = 0.8)+# by changing alpha, we change the transparency
  geom_smooth(span = 0.15, se = FALSE)+# by changing span, I control the wiggles in the trend
  ylab("Unit discharge (cm)") +
  xlab("Water day")+
  scale_color_manual(values = c("#00BFC4","#F8766D"))+
  scale_fill_manual(values = c("#00BFC4","#F8766D"))+
  theme_fg +
  facet_wrap(~ws.f)
uq_plot
```


Another look at these data from a frequency/probabilistic perspective:

```{r}
uq_p_plot <- ggplot(hjg, aes(x=uq.cm,color= ws.f, fill = ws.f))+
  geom_density(alpha = 0.5, size = 1.0)+
  ylab("Kernel Density Estimate") +
  xlab("Unit discharge (cm)")+
  scale_color_manual(values = c("#00BFC4","#F8766D"))+
  scale_fill_manual(values = c("#00BFC4","#F8766D"))+
  theme_fg +
  facet_wrap(~ws.f)
uq_p_plot
```

**Question**: Do these distributions look normal?

==========================N E X T  S E S S I O N========================================

Let's try to use a different scale for the x axis:


```{r}
uq_d_plot  <- uq_p_plot + scale_x_log10() + theme(legend.position = c(0.1,0.8))
uq_d_plot
```

**Question**: Does it look like the distribution of the streamflow is the same?

Let's take a look at the seasonal changes to explore more about the potential differences:

```{r}
ssn_plot <- ggplot(hjg, aes(x=uq.cm, color=as.factor(ssn), fill = as.factor (ssn)))+
  geom_density(alpha = 0.5)+
  ylab("Kernel density estimate") +
  xlab("Unit discharge (cm)")+
  scale_x_log10()+
  theme_fg +
  theme(legend.position = c(0.1,0.8))+
  facet_wrap(~ws.f)
ssn_plot
```

Now, lets compare both watersheds within each season

```{r}
ssn_plot_c <- ggplot(hjg, aes(x=uq.cm, color=as.factor(ws.f), fill = as.factor (ws.f)))+
  geom_density(alpha = 0.2, size = 1.2)+
  ylab("Kernel density estimate") +
  xlab("Unit discharge (cm)")+
  scale_x_log10()+
  scale_color_manual(values = c("#00BFC4","#F8766D"))+
  scale_fill_manual(values = c("#00BFC4","#F8766D"))+
  theme_fg +
  theme(legend.position = c(0.1,0.8))+
  facet_wrap(~ssn, nrow = 2)
ssn_plot_c
```
==========================N E X T  S E S S I O N========================================


## Your homework:

-   How would you explain the concept of unit discharge and its measurement to a lay person?
-   What would you say are the differences in looking at discharge data as a time series or as a histogram/density plot?
-   What is the difference between a Normal distribution and a log-normal distribution.

### For our next homework:
-   Why do you think discharge data tend to fit a log-normal distribution. Explain
